---
title: 'Reproducible Research - Course Project 1 - Activity Monitoring Data'
author: "Santiago Ramirez"
date: "15 November 2017"
output: html_document

---

##Dataset Description
The variables included in the dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

```{r setup, include=FALSE}
rm(list = ls())
library(knitr)
library(lattice)
library(ggplot2)
library(dplyr)
opts_chunk$set(echo = TRUE, cache = TRUE) ## Always show answers and avoid repeating unnecessary calculations
```


##Loading and preprocessing the data

**1. - Load the data**

```{r readingpraparingdata}
activity <- read.csv("activity.csv", sep = ",")
head(activity)

#To know about the data
str(activity)
summary(activity)
```

**2. - Process/transform the data (if necessary) into a format suitable for your analysis **

```{r processingdata}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
#validate if type of data for "Date" has changed
str(activity)
```
Now, date is of type "Date"

##What is mean total number of steps taken per day?

**1. - Calculate the total number of steps taken per day **

``` {r meansteps}
total.steps <- tapply(activity$steps, activity$date, FUN = sum, na.rm = TRUE)
head(total.steps)
```

**2. - Make a histogram of the total number of steps taken each day **

```{r histogramsteps, fig.height=4}
hist(total.steps, xlab = "day", ylab = "Total number of steps", main = "Total number of steps taken each day", col = "cyan")
```

**3. - Calculate and report the mean and median of the total number of steps taken per day **

```{r MeanMedianStepsDay}
#First calculate the mean
mean.steps <- tapply(activity$steps, activity$date, FUN = mean, na.rm = TRUE)
#Second calcuate the median
median.steps <- tapply(activity$steps, activity$date, FUN = median, na.rm = TRUE)

print(mean.steps)
print(median.steps)
```

##What is the average daily activity pattern?

**1. - Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r averagedaily}
averages.steps <- aggregate(x = list(steps = activity$steps), by = list(interval = activity$interval), 
    FUN = mean, na.rm = TRUE)
ggplot(data = averages.steps, aes(x = interval, y = steps)) + geom_line() + xlab("5-minute interval") + 
    ylab("average number of steps") + ggtitle("Average number of steps per time Interval")
```

**2. - Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**
```{r averagesmax}
averages.steps[which.max(averages.steps$steps),]
```

The 5-minute interval with the (average) maximum number of steps is the interval **835** with **206** steps on average.


##Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA).
The presence of missing days may introduce bias into some calculations or summaries of the data.

**1. - Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r totalmissingvalues}

#Since it might be possible that one observation could have more that 1 NA, the count will be done by obsrevations,
#regardless the amount of NA present on each observation, the fact of finding 1, 2 or 3 NA, it will be considered as 1

dimension <- dim(activity)[1]
counter <- 1
nacounter <- 0
nasum <- 0

while (counter <= dimension) {
        nasum <- sum(is.na(activity[counter,]))
        if (nasum > 0) {
                nacounter <- nacounter + 1
                nasum <- 0
        }
        counter <- counter + 1
}

print(nacounter)

# To compare, calculate the total amount of NA in the data set

sum(is.na(activity))
```

The total amount of missing values in the observations is **`r nacounter`**.
Important to highlight that is the same value of NA present, which indicate that each observation
has not more than 1 NA value.

**2. - Devise a strategy for filling in all of the missing values in the dataset. The  strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

First, create a table with the average steps per day

```{r avgstepsDay}
averages.stepsDay <- aggregate(x = list(stepsAvg = activity$steps), by = list(date = activity$date), FUN = mean, na.rm = TRUE)

head(averages.stepsDay, 20)
```

Some days have no steps information. This fields will be updated with the general steps average

``` {r generalstepsaverage}
avgsteps <- mean(activity$steps, na.rm = TRUE)
```

Fill the gaps with avgsteps

```{r fillinggeneralgaps}
dimension <- dim(averages.stepsDay)[1]
index <- 1
while (index <= dimension) {
        if (is.na(averages.stepsDay[index, 2])) {
                averages.stepsDay[index, 2] <- avgsteps
        }
        index <- index + 1
}

head(averages.stepsDay, 20)
```

There is now a value for each day.

Next step is to add the corresponding step value in the orignal table, according to the date


``` {r newactivitilist}
activity2 <- merge(x = activity, y = averages.stepsDay, by.x = "date")

#If data in "steps" is NA, it will be replaced by the value in "stepAvg"
dimension <- dim(activity2)[1]
index <- 1
while (index <= dimension) {
        if (is.na(activity2[index, 2])) {
                activity2[index, 2] <- activity2[index, 4]
        }
        index <- index + 1
}

```



 







